# Model Comparison, Partial F Tests, and ANOVA {#sec-model-comparison}

In this chapter, we study the ANOVA decomposition for linear regression, the overall significance test, and the comparison of nested models. These ideas connect the geometry of least squares with the inferential tools developed in earlier chapters.

> **Learning Objectives**
>
> By the end of this chapter, students should be able to:
>
> -   define the total, regression, and error sums of squares;
> -   explain the ANOVA decomposition in regression with an intercept;
> -   interpret the degrees of freedom associated with SST, SSR, and SSE;
> -   perform the overall $F$ test for regression;
> -   compare nested linear models using extra sums of squares;
> -   interpret ANOVA tables produced by statistical software.

**Reading**

Recommended reading for this chapter:

-   Seber and Lee:
    -   sections on analysis of variance in regression
    -   sums of squares and decomposition of variability
    -   tests for nested models
-   Montgomery, Peck, and Vining:
    -   sections on the ANOVA table
    -   overall model significance
    -   partial and sequential sums of squares

**Review of the Linear Model**

Recall the normal linear model

$$
\mathbf{Y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\varepsilon},
\qquad
\boldsymbol{\varepsilon} \sim N_n(\mathbf{0}, \sigma^2 \mathbf{I}_n).
$$

When $\mathbf{X}$ has full column rank, the ordinary least squares estimator is

$$
\hat{\boldsymbol{\beta}}
=
(\mathbf{X}^\top \mathbf{X})^{-1}\mathbf{X}^\top \mathbf{Y},
$$

and the fitted values are

$$
\hat{\mathbf{Y}} = \mathbf{X}\hat{\boldsymbol{\beta}}.
$$

The residual vector is

$$
\mathbf{e} = \mathbf{Y} - \hat{\mathbf{Y}}.
$$

From Chapter 2, we know that $\hat{\mathbf{Y}}$ and $\mathbf{e}$ are orthogonal. From Chapter 3, we know that this leads to useful distributional results for inference. In this chapter, we organize these ideas into the ANOVA framework.

## Total, Explained, and Unexplained Variation

A central question in regression is: How much of the variation in the response can be explained by the model? To answer this, we decompose the total variation in $\mathbf{Y}$ into:

-   variation explained by regression;
-   variation left unexplained by the model.

When the model includes an intercept, this decomposition takes a particularly simple and important form.

::: {#def-regression-sums-of-squares}

## Regression Sums of Squares

For an OLS fit with an intercept, let $\bar Y=n^{-1}\sum_{i=1}^nY_i$, $\hat{\mathbf Y}=\mathbf X\hat{\boldsymbol\beta}$, and $\mathbf e=\mathbf Y-\hat{\mathbf Y}$. Define the **total**, **error**, and **regression sums of squares** by

$$
\begin{aligned}
\mathrm{SST}&=\sum_{i=1}^n(Y_i-\bar Y)^2
=\|\mathbf Y-\bar Y\mathbf 1\|^2,\\
\mathrm{SSE}&=\sum_{i=1}^n(Y_i-\hat Y_i)^2
=\mathbf e^\top\mathbf e,\\
\mathrm{SSR}&=\sum_{i=1}^n(\hat Y_i-\bar Y)^2
=\|\hat{\mathbf Y}-\bar Y\mathbf 1\|^2.
\end{aligned}
$$

:::

SST measures variation around the sample mean; SSE measures deviations from the fitted model; SSR measures how much the fitted values vary around the sample mean. These are definitions. Their additive relationship below follows from OLS orthogonality and the inclusion of an intercept.

::: {#thm-anova-decomposition}

## ANOVA Decomposition with an Intercept

For an ordinary least squares fit whose design matrix contains an intercept,

$$
\mathrm{SST}=\mathrm{SSR}+\mathrm{SSE}.
$$

The centered fitted vector $\hat{\mathbf Y}-\bar Y\mathbf 1$ is orthogonal to the residual vector $\mathbf e$. Thus total variation around the sample mean splits into explained variation and residual variation. This geometric identity does not require normal errors.

:::

**Why the Decomposition Holds**

The key reason is orthogonality. We can write

$$
\mathbf{Y} - \bar{Y}\mathbf{1}
=
(\hat{\mathbf{Y}} - \bar{Y}\mathbf{1})
+
(\mathbf{Y} - \hat{\mathbf{Y}}).
$$

That is,

$$
\mathbf{Y} - \bar{Y}\mathbf{1}
=
(\hat{\mathbf{Y}} - \bar{Y}\mathbf{1}) + \mathbf{e}.
$$

Because the model contains an intercept, the vector $\bar{Y}\mathbf{1}$ lies in the column space of $\mathbf{X}$. Hence both $\hat{\mathbf{Y}}$ and $\bar{Y}\mathbf{1}$ lie in the model space, so their difference also lies in the model space. Since the residual vector $\mathbf{e}$ is orthogonal to the model space, we have

$$
(\hat{\mathbf{Y}} - \bar{Y}\mathbf{1})^\top \mathbf{e} = 0.
$$

Therefore, by the Pythagorean theorem,

$$
\|\mathbf{Y} - \bar{Y}\mathbf{1}\|^2
=
\|\hat{\mathbf{Y}} - \bar{Y}\mathbf{1}\|^2
+
\|\mathbf{e}\|^2,
$$

which is exactly

$$
\mathrm{SST} = \mathrm{SSR} + \mathrm{SSE}.
$$

**Degrees of Freedom**

The ANOVA decomposition is accompanied by a decomposition of degrees of freedom. When the model includes an intercept and $\mathbf{X}$ has rank $p$, we have:

-   total degrees of freedom: $n-1$;
-   regression degrees of freedom: $p-1$;
-   error degrees of freedom: $n-p$.

Thus,

$$
n-1 = (p-1) + (n-p).
$$

These match the sum of squares decomposition:

$$
\mathrm{SST} = \mathrm{SSR} + \mathrm{SSE}.
$$

::: {#def-regression-mean-squares}

## Regression and Error Mean Squares

A **mean square** is a sum of squares divided by its associated degrees of freedom. For a full-rank regression design with an intercept and $1<p<n$, define

$$
\mathrm{MSR}=\frac{\mathrm{SSR}}{p-1},\qquad
\mathrm{MSE}=\frac{\mathrm{SSE}}{n-p}.
$$

Here $p$ includes the intercept coefficient.

:::

Dividing by degrees of freedom accounts for the different dimensions contributing to each sum of squares. Under the linear-model moment assumptions from Chapter 3, MSE is an unbiased estimator of $\sigma^2$. MSR estimates the same noise level under the overall null hypothesis, which motivates their ratio.

## The Overall F Test

A major inferential question is whether the regression model provides any explanatory power beyond the intercept-only model. Suppose the model includes an intercept and $p-1$ additional predictors. The null hypothesis is

$$
H_0: \beta_1 = \beta_2 = \cdots = \beta_{p-1} = 0,
$$

where $\beta_0$ is the intercept and $p$ counts all regression coefficients. Equivalently, under $H_0$, the mean response does not depend on the predictors. The alternative is that at least one non-intercept coefficient is nonzero.

::: {#thm-overall-f}

## Overall F Test

Under the normal linear model with a fixed, full-rank design containing an intercept and $1<p<n$, the null hypothesis that all non-intercept coefficients are zero gives

$$
F=\frac{\mathrm{SSR}/(p-1)}{\mathrm{SSE}/(n-p)}
=\frac{\mathrm{MSR}}{\mathrm{MSE}}
\sim F_{p-1,n-p}.
$$

Large values provide evidence against the intercept-only mean model. Exact finite-sample inference requires the normal, common-variance error assumptions; the sum-of-squares identity alone does not supply this distribution.

:::

**Interpretation of the Overall F Test**

The numerator measures explained variation per regression degree of freedom. The denominator measures unexplained variation per residual degree of freedom. So the $F$ statistic compares:

-   how much signal the model explains;
-   how much noise remains in the residuals.

If the predictors have no effect, then both quantities should be of similar size, and the ratio should not be unusually large. If the predictors explain substantial variation, then the numerator should be much larger than the denominator.

**Relationship to the Intercept-Only Model**

The overall $F$ test compares two models:

-   the reduced model: intercept only;
-   the full model: intercept plus predictors.

Thus the ANOVA decomposition provides the basis for formal model comparison. This leads naturally to the idea of nested models.

## Nested Models

::: {#def-nested-linear-models}

## Nested Linear Models

For the same response vector and observations, let the reduced and full models have mean spaces $\mathcal C(\mathbf X_R)$ and $\mathcal C(\mathbf X_F)$. The models are **nested** when

$$
\mathcal C(\mathbf X_R)\subseteq\mathcal C(\mathbf X_F).
$$

Thus every mean vector allowed by the reduced model is also allowed by the full model. A strict comparison requires ranks $p_R<p_F$; adding columns that are linear combinations of existing columns does not enlarge the mean space.

:::

::: {#exm-nested-regression-models}

## Adding a Predictor

Consider the reduced model $Y_i=\beta_0+\beta_1x_{i1}+\varepsilon_i$ and the full model $Y_i=\beta_0+\beta_1x_{i1}+\beta_2x_{i2}+\varepsilon_i$. Setting $\beta_2=0$ recovers the reduced model. If the $x_2$ column is not in the span of the intercept and $x_1$, the full model adds one degree of freedom.

:::

::: {#def-extra-sum-of-squares}

## Extra Sum of Squares

For nested models fitted by OLS to the same response and observations, define the **extra sum of squares** for the full model relative to the reduced model as

$$
\mathrm{SS}_{\mathrm{extra}}=\mathrm{SSE}_R-\mathrm{SSE}_F.
$$

Its associated degrees of freedom are $p_F-p_R$, where $p_R$ and $p_F$ are the design-matrix ranks.

:::

The full model minimizes residual variation over a larger mean space, so $\mathrm{SSE}_F\le\mathrm{SSE}_R$. Extra sum of squares therefore measures a nonnegative improvement in fit. A positive improvement alone is not evidence against the reduced model: the F test scales that improvement by its degrees of freedom and the estimated noise variance.

::: {#thm-partial-f}

## Partial F Test for Nested Normal Linear Models

Fit both models to the **same response and the same observations**. Suppose their fixed design matrices have nested column spaces, with ranks $p_R<p_F<n$, and the full model has independent $N(0,\sigma^2)$ errors. Under the null hypothesis that the mean belongs to the reduced model,

$$
F=\frac{(\mathrm{SSE}_R-\mathrm{SSE}_F)/(p_F-p_R)}{\mathrm{SSE}_F/(n-p_F)}
\sim F_{p_F-p_R,n-p_F}.
$$

Large values give evidence against the reduced mean model. The numerator measures improvement per added degree of freedom; the denominator estimates the unexplained variance under the full model.

:::

::: {.callout-important title="Compare the models, not just the printed p-values"}

Different missing-value patterns can cause two R fits to use different observations. Construct a common analysis dataset before comparing them. For one added coefficient, the partial F statistic equals the square of its two-sided t-test statistic from the full model.

:::

**Connection with General Linear Hypotheses**

Nested column-space models correspond to linear restrictions on the full-model coefficients. More generally, Chapter 3 allows a specified affine restriction of the form

$$
H_0: \mathbf{C}\boldsymbol{\beta} = \mathbf{d}.
$$

So the ANOVA comparison of nested models is another way of expressing the general $F$ test from Chapter 3. In practice, this is one of the most common uses of regression ANOVA tables.

::: {#def-sequential-partial-sums-of-squares}

## Sequential and Partial Sums of Squares

For a term in a specified regression model, both quantities are extra sums of squares from a nested comparison on the same observations:

- **Sequential sum of squares:** the reduction in SSE when the term is added after the preceding terms in a specified order.
- **Partial sum of squares:** the increase in SSE when the specified term is removed from a full model while the other specified terms are retained.

:::

When predictors are correlated, sequential sums of squares can depend on their order. Partial sums of squares answer a comparison conditional on the terms retained in the reduced model. With interactions or rank deficiency, state the exact restriction and respect model hierarchy rather than relying only on a Type I, II, or III label.

::: {#def-coefficient-of-determination}

## Coefficient of Determination

For an OLS fit with an intercept and $\mathrm{SST}>0$, the **coefficient of determination** is

$$
R^2=\frac{\mathrm{SSR}}{\mathrm{SST}}
=1-\frac{\mathrm{SSE}}{\mathrm{SST}}.
$$

:::

By the ANOVA decomposition, $R^2$ lies between 0 and 1 and measures the proportion of centered variation explained by the fitted model. If every observed response is identical, $\mathrm{SST}=0$ and this ratio is undefined.

**Interpretation of R Squared**

-   If $R^2$ is close to 1, the model explains a large proportion of the variation in the response.
-   If $R^2$ is close to 0, the model explains little of the variation.

However, $R^2$ alone does not guarantee that the model is appropriate. A high $R^2$ does not ensure that assumptions are satisfied, and a low $R^2$ does not necessarily imply the model is useless.

::: {#def-adjusted-r-squared}

## Adjusted Coefficient of Determination

For an OLS fit with an intercept, $\mathrm{SST}>0$, and design rank $p<n$, define the **adjusted coefficient of determination** as

$$
R^2_{\mathrm{adj}}
=1-\frac{\mathrm{SSE}/(n-p)}{\mathrm{SST}/(n-1)}.
$$

:::

Ordinary $R^2$ cannot decrease when the model space is enlarged on the same data. Adjusted $R^2$ accounts for the loss of residual degrees of freedom and can decrease or become negative. It compares residual mean square with the sample variance of the response; it is not a literal proportion of explained variation or an estimate of out-of-sample predictive performance.

::: {#exm-anova-by-hand}

## A Small ANOVA Calculation

For the column vectors $\mathbf x=(0,1,2,3)$ and $\mathbf y=(1,3,3,5)$, the fitted line is $\hat y=1.2+1.2x$. Therefore $\hat{\mathbf y}=(1.2,2.4,3.6,4.8)$, $\mathbf e=(-0.2,0.6,-0.6,0.2)$, and $\bar y=3$.

$$
\begin{aligned}
\mathrm{SST}&=(1-3)^2+(3-3)^2+(3-3)^2+(5-3)^2=8,\\
\mathrm{SSE}&=(-0.2)^2+0.6^2+(-0.6)^2+0.2^2=0.8,\\
\mathrm{SSR}&=\mathrm{SST}-\mathrm{SSE}=8-0.8=7.2.
\end{aligned}
$$

Here $n=4$ and $p=2$, so the regression, residual, and total degrees of freedom are 1, 2, and 3, respectively.

| Source | Sum of squares | df | Mean square | F |
|---|---:|---:|---:|---:|
| Regression | 7.2 | 1 | 7.2 | 18 |
| Error | 0.8 | 2 | 0.4 | |
| Total | 8.0 | 3 | | |

The model has $R^2=7.2/8=0.9$, yet the F-test p-value is $P(F_{1,2}\ge18)\approx0.0513$. With only two residual degrees of freedom, a high training $R^2$ does not automatically give evidence at the 5% significance level.

:::

**ANOVA Table Structure**

A standard regression ANOVA table has the following structure:

| Source     | Sum of Squares | Degrees of Freedom | Mean Square | F       |
|------------|----------------|--------------------|-------------|---------|
| Regression | SSR            | $p-1$              | MSR         | MSR/MSE |
| Error      | SSE            | $n-p$              | MSE         |         |
| Total      | SST            | $n-1$              |             |         |

Students should learn to move fluently between:

-   formulas;
-   geometric interpretation;
-   software output.

## R Demonstration

**Fit a simple regression model**

```{r}
x <- c(0, 1, 2, 3)
y <- c(1, 3, 3, 5)

fit <- lm(y ~ x)
summary(fit)
```

**Obtain the ANOVA table**

```{r}
anova(fit)
```

**Verify sums of squares manually**

```{r}
ybar <- mean(y)
yhat <- fitted(fit)
e <- resid(fit)

SST <- sum((y - ybar)^2)
SSE <- sum(e^2)
SSR <- sum((yhat - ybar)^2)

c(SST = SST, SSR = SSR, SSE = SSE)
SST - SSR - SSE
```

**See the three sources of variation.** Each vertical segment below represents one deviation. Squaring and adding the four segment lengths gives the corresponding sum of squares.

```{r}
#| label: fig-anova-variation
#| fig-cap: "Total deviations from the sample mean, explained deviations of fitted values from the mean, and residual deviations from the fitted line. SST = SSR + SSE holds after summing across observations; it is not a point-by-point identity between squared deviations."
#| fig-width: 7
#| fig-height: 3.2
par(mfrow = c(1, 3), mar = c(4, 4, 2.4, 0.6),
    mgp = c(2.4, 0.7, 0))
for (component in c("Total", "Explained", "Residual")) {
  plot(x, y, pch = 16, ylim = c(0.5, 5.5), xlab = "x", ylab = "y",
       main = component, cex.main = 1)
  abline(h = ybar, col = "gray60", lty = 2)
  abline(fit, col = "steelblue4")
  if (component == "Total") {
    segments(x, ybar, x, y, col = "gray35", lwd = 2)
  } else if (component == "Explained") {
    segments(x, ybar, x, yhat, col = "steelblue4", lwd = 2)
    points(x, yhat, pch = 1, col = "steelblue4")
  } else {
    segments(x, yhat, x, y, col = "darkorange3", lwd = 2)
  }
}
par(mfrow = c(1, 1))
```

**Compute R squared manually**

```{r}
R2 <- SSR / SST
R2
summary(fit)$r.squared
summary(fit)$adj.r.squared
pf(18, df1 = 1, df2 = 2, lower.tail = FALSE)
```

**Compare nested models**

```{r}
dat <- data.frame(
  y = c(4, 5, 7, 10, 8, 12, 13, 14),
  x1 = c(1, 2, 3, 4, 5, 6, 7, 8),
  x2 = c(2, 1, 3, 2, 5, 4, 6, 5)
)

fit_reduced <- lm(y ~ x1, data = dat)
fit_full <- lm(y ~ x1 + x2, data = dat)

anova(fit_reduced, fit_full)
```

**Inspect the two fitted models**

```{r}
summary(fit_reduced)
summary(fit_full)
```

**Interpretation of Software Output**

For a fitted model in R:

-   `anova(fit)` gives sequential sums of squares in the order terms enter a single model;
-   `anova(fit_reduced, fit_full)` compares nested models;
-   `summary(fit)` reports the overall $F$ statistic, $R^2$, and adjusted $R^2$.

Students should understand that these outputs are not separate topics. They are all built from the same least squares geometry and distribution theory.

**In-Class Discussion Questions**

1.  Why does the ANOVA decomposition require an intercept for the usual SST = SSR + SSE identity?
2.  Why must $\mathrm{SSE}_F \le \mathrm{SSE}_R$ for nested models?
3.  What does the overall $F$ test tell us that individual $t$ tests do not?
4.  Why can $R^2$ be misleading if used alone?

## Practice Problems

**Conceptual**

1.  Explain the meaning of SST, SSR, and SSE in words.
2.  Explain why the regression degrees of freedom are $p-1$ when the model includes an intercept.
3.  Explain the difference between the overall $F$ test and a test for a single coefficient.

**Computational**

Suppose a regression model with intercept has:

-   $n=20$,
-   $p=4$,
-   $\mathrm{SST}=100$,
-   $\mathrm{SSE}=40$.

Compute:

1.  $\mathrm{SSR}$,
2.  the degrees of freedom for regression and error,
3.  $\mathrm{MSR}$,
4.  $\mathrm{MSE}$,
5.  the overall $F$ statistic,
6.  $R^2$.

**Nested Model Problem**

A reduced model has $\mathrm{SSE}_R = 120$ with $p_R = 3$, and a full model has $\mathrm{SSE}_F = 90$ with $p_F = 5$. If $n=30$, compute the nested-model $F$ statistic.

**Suggested Homework**

Complete the following tasks:

-   prove the decomposition $\mathrm{SST}=\mathrm{SSR}+\mathrm{SSE}$ when the model includes an intercept;
-   derive the overall $F$ statistic from the ANOVA decomposition;
-   fit a regression model in R and reproduce the ANOVA table by hand;
-   compare two nested models using an extra sum of squares test;
-   interpret both $R^2$ and adjusted $R^2$ for a chosen dataset.

**Summary**

In this chapter, we developed the ANOVA framework for linear regression. We defined:

$$
\mathrm{SST}, \qquad \mathrm{SSR}, \qquad \mathrm{SSE},
$$

and showed that, with an intercept,

$$
\mathrm{SST} = \mathrm{SSR} + \mathrm{SSE}.
$$

This decomposition led to:

-   the ANOVA table;
-   the overall $F$ test for regression;
-   the comparison of nested models through extra sums of squares;
-   the interpretation of $R^2$ and adjusted $R^2$.
