# Model Assessment and Prediction {#sec-model-assessment}

A useful model must answer a stated question. Estimating an adjusted association, comparing treatments, and predicting a future response can favor different choices. This chapter keeps the main principles of collinearity and validation; automatic selection algorithms are in an optional appendix.

## Stable estimates and scientific questions

**Collinearity.** When columns of $X$ are nearly dependent, different coefficient combinations can produce very similar fitted values. The covariance $\sigma^2(X^\top X)^{-1}$ then explains why individual coefficients can be unstable even when fitted values are reasonably precise.

For a single nonconstant predictor column, $\operatorname{VIF}_j=1/(1-R_j^2)$, where $R_j^2$ comes from regressing that column on the other predictors with an intercept. A large value is a diagnostic clue, not an automatic rule for deleting variables. Multi-column factors require additional care in interpreting column-specific VIFs.

**Keep the question fixed.** Removing a covariate changes the adjusted comparison. Retain terms required by the scientific question or design, and respect hierarchy when including interactions. Centering can clarify an intercept or reduce some polynomial-column correlations, but it does not create new information.

::: {.callout-important title="Selection changes inference"}

Intervals and tests derived for a prespecified model do not automatically retain their interpretation after searching the data for a favorable model. A training/test split or cross-validation estimates predictive performance; it does not by itself validate post-selection coefficient intervals.

:::

## Comparing predictive performance

Training SSE and training $R^2$ reward added flexibility. Predictive evaluation instead asks how a fitted procedure behaves on observations it did not use for fitting.

**A simple workflow.** Specify a small set of candidate models, use the same observations and the same folds, fit each model only on the training portion, and evaluate it on the held-out portion. Any imputation, centering, feature construction, or variable search learned from data belongs inside the training step. If validation is used to choose a model, its winning validation error is itself optimistic; use an untouched test set or nested resampling for a final performance claim.

For continuous responses, one useful loss is $\operatorname{MSE}_{test}=m^{-1}\sum_{i=1}^m(y_i-\hat y_i)^2$. For binary and count outcomes, the GLM chapter introduces response-appropriate interpretation and predictive checks.

::: {.callout-note title="Respect the observational unit"}

Random row-wise folds are appropriate only when they match the intended prediction setting. With repeated measurements, keep a subject's observations together; with temporal prediction, evaluate on later periods. A large number of rows is not a substitute for independent subjects or future time periods.

:::

**AIC and model comparison.** AIC balances likelihood fit and parameter count. It is useful when candidate likelihoods refer to the same response and comparable data, but it is not an effect estimate or a significance test. Comparing AIC values from a model for $Y$ and one for $\log Y$ requires accounting for the transformation of the likelihood; do not compare the raw software values mechanically. BIC, $C_p$, and automatic searches are in [Automatic Model Selection (Optional)](App_C-model-selection.qmd).

## A small validation example

The following simulated curvature example uses the same five folds for two prespecified mean models. It illustrates predictive evaluation, not a guarantee that a quadratic model will always win.

```{r}
set.seed(8561)
n <- 100
x <- runif(n, -2, 2)
dat <- data.frame(x = x, y = 2 + x + 0.7 * x^2 + rnorm(n, sd = 1.5))
fold <- sample(rep(1:5, length.out = n))
models <- list(linear = y ~ x, quadratic = y ~ x + I(x^2))
pred <- matrix(NA_real_, n, length(models),
               dimnames = list(NULL, names(models)))
for (k in 1:5) {
  for (j in seq_along(models)) {
    fit <- lm(models[[j]], data = dat[fold != k, ])
    pred[fold == k, j] <- predict(fit, newdata = dat[fold == k, ])
  }
}
cv_mse <- colMeans((dat$y - pred)^2)
cv_mse
```

```{r}
#| label: fig-validation-predictions
#| fig-cap: "Predictions made for held-out observations. Every plotted prediction comes from a fit that excluded that observation's fold."
#| fig-width: 7
#| fig-height: 3.2
par(mfrow = c(1, 2))
for (j in seq_along(models)) {
  plot(dat$y, pred[, j], pch = 19, col = adjustcolor("steelblue", 0.6),
       xlab = "Observed response", ylab = "Held-out prediction",
       main = names(models)[j])
  abline(0, 1, lty = 2)
}
par(mfrow = c(1, 1))
```

**Report more than a winner.** State the loss, the splitting rule, the population or future setting of interest, and how much performance differs. Investigate poor predictions. A single difference in estimated MSE does not establish a universal ranking of the methods.

## Practice

1. Explain why adding a predictor can improve training fit without improving held-out predictions.
2. Explain why a confounder or design variable should not be removed solely because of its p-value or VIF.
3. Adapt the example to grouped data by assigning folds at the group level.
4. Write a short report distinguishing an inferential conclusion about a coefficient from a predictive conclusion about a fitted procedure.
