---
show-solutions: true
---

# Assignment 1 {.unnumbered}

**STAT 8561 \| Fall 2026**\
**Total:** 40 points\
**Coverage:** [Chapter 1 — Matrix Representation and Notation](01_matrix_representation.qmd) and [Chapter 2 — Least Squares Estimation](02_LSE.qmd).

This assignment focuses on matrix representation, least squares estimation, projection geometry, and linear regression in R.

::: {tbl-colwidths="[16,72,12]"}

| Question | Topic | Points |
|:---:|:---:|:---:|
| 1 | Matrix representation | 8 |
| 2 | Least squares and projection | 12 |
| 3 | Consequences and geometry of least squares | 12 |
| 4 | Regression in R | 8 |
| **Total** |  | **40** |

:::

::: {.callout-note title="Instructions"}

Use **Quarto** or **R Markdown**. Upload **both** files to the **Assignment 1** submission folder on **iCollege**:

1. Your rendered **PDF**, named `A1_LastName_FirstName.pdf`.
2. Its matching source file, named `A1_LastName_FirstName.qmd` or `A1_LastName_FirstName.Rmd`.

The PDF must be generated from the submitted source file. **HTML alone does not satisfy the submission requirement.**

- Put your name and course number at the beginning of the PDF.
- Clearly label every question and subpart.
- For each multiple-choice item, select **exactly one** answer and report its letter. Include an explanation only when requested.
- For True/False items, enter only `T` or `F`.
- For short-answer questions, one or two clear sentences are sufficient unless a derivation is requested.
- Round numerical answers to **three decimal places**, retaining full precision during calculations.
- Include all R code used to obtain your answers in executable chunks.
- Your source must render from beginning to end without errors. Type mathematical expressions using LaTeX; do not submit screenshots of answers or code.

:::

## Question 1: Matrix Representation of the Linear Model [8]

Consider the simple linear regression model

$$
Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i,
\qquad i=1,\ldots,5,
$$

with the following observed data:

| $i$ | $x_i$ | $Y_i$ |
|---:|---:|---:|
| 1 | 0 | 2 |
| 2 | 1 | 4 |
| 3 | 2 | 5 |
| 4 | 3 | 8 |
| 5 | 4 | 9 |

### (a) [2]

Which of the following is the correct design matrix $\mathbf{X}$?

A.  $$
    \begin{pmatrix}
    0 & 2\\
    1 & 4\\
    2 & 5\\
    3 & 8\\
    4 & 9
    \end{pmatrix}
    $$

B.  $$
    \begin{pmatrix}
    1 & 0\\
    1 & 1\\
    1 & 2\\
    1 & 3\\
    1 & 4
    \end{pmatrix}
    $$

C.  $$
    \begin{pmatrix}
    0 & 1\\
    1 & 1\\
    2 & 1\\
    3 & 1\\
    4 & 1
    \end{pmatrix}
    $$

D.  $$
    \begin{pmatrix}
    1 & 2\\
    1 & 4\\
    1 & 5\\
    1 & 8\\
    1 & 9
    \end{pmatrix}
    $$

:::: {.content-visible when-meta="show-solutions"}

::: {#a1-solution-1a .callout-tip collapse="true" title="Solution 1(a)"}

**B.** Each row is $(1,x_i)$: the first column represents the intercept, and the second contains the predictor values.

:::

::::

### (b) [2]

Fill in the following table.

| Quantity | Dimension / Rank |
|:---|:---|
| $\mathbf{Y}$ | `___` |
| $\mathbf{X}$ | `___` |
| $\boldsymbol{\beta}$ | `___` |
| $\operatorname{rank}(\mathbf{X})$ | `___` |

:::: {.content-visible when-meta="show-solutions"}

::: {#a1-solution-1b .callout-tip collapse="true" title="Solution 1(b)"}

| Quantity | Dimension / Rank |
|:---|:---|
| $\mathbf{Y}$ | $5\times 1$ |
| $\mathbf{X}$ | $5\times 2$ |
| $\boldsymbol{\beta}$ | $2\times 1$ |
| $\operatorname{rank}(\mathbf{X})$ | $2$ |

The two columns of $\mathbf{X}$ are linearly independent because the predictor is not constant.

:::

::::

### (c) [2]

Which pair is correct?

A.  $$
    \mathbf{X}^\top\mathbf{X}
    =
    \begin{pmatrix}
    5 & 10\\
    10 & 30
    \end{pmatrix},
    \qquad
    \mathbf{X}^\top\mathbf{Y}
    =
    \begin{pmatrix}
    28\\
    74
    \end{pmatrix}
    $$

B.  $$
    \mathbf{X}^\top\mathbf{X}
    =
    \begin{pmatrix}
    5 & 5\\
    5 & 30
    \end{pmatrix},
    \qquad
    \mathbf{X}^\top\mathbf{Y}
    =
    \begin{pmatrix}
    28\\
    74
    \end{pmatrix}
    $$

C.  $$
    \mathbf{X}^\top\mathbf{X}
    =
    \begin{pmatrix}
    30 & 10\\
    10 & 5
    \end{pmatrix},
    \qquad
    \mathbf{X}^\top\mathbf{Y}
    =
    \begin{pmatrix}
    77\\
    28
    \end{pmatrix}
    $$

D.  $$
    \mathbf{X}^\top\mathbf{X}
    =
    \begin{pmatrix}
    5 & 10\\
    10 & 20
    \end{pmatrix},
    \qquad
    \mathbf{X}^\top\mathbf{Y}
    =
    \begin{pmatrix}
    28\\
    77
    \end{pmatrix}
    $$

:::: {.content-visible when-meta="show-solutions"}

::: {#a1-solution-1c .callout-tip collapse="true" title="Solution 1(c)"}

**A.** Here $\sum_i x_i=10$, $\sum_i x_i^2=30$, $\sum_i Y_i=28$, and $\sum_i x_iY_i=74$. Thus

$$
\mathbf{X}^\top\mathbf{X}=\begin{pmatrix}5&10\\10&30\end{pmatrix},
\qquad
\mathbf{X}^\top\mathbf{Y}=\begin{pmatrix}28\\74\end{pmatrix}.
$$

:::

::::

### (d) [2]

Assume

$$
\mathbb{E}[\boldsymbol{\varepsilon}] = \mathbf{0},
\qquad
\operatorname{Var}(\boldsymbol{\varepsilon})=\sigma^2\mathbf{I}_5.
$$

For each statement, enter `T` or `F`.

::: {tbl-colwidths="[88,12]"}

| Statement | T/F |
|:---|:---:|
| $\mathbb{E}[\mathbf{Y}] = \mathbf{X}\boldsymbol{\beta}$ | `___` |
| $\operatorname{Var}(\mathbf{Y}) = \sigma^2\mathbf{I}_5$ | `___` |
| $\mathbb{E}[\mathbf{Y}] = \mathbf{0}$ | `___` |
| The five errors all have variance $\sigma^2$ | `___` |

:::

:::: {.content-visible when-meta="show-solutions"}

::: {#a1-solution-1d .callout-tip collapse="true" title="Solution 1(d)"}

::: {tbl-colwidths="[88,12]"}

| Statement | T/F |
|:---|:---:|
| $\mathbb{E}[\mathbf{Y}] = \mathbf{X}\boldsymbol{\beta}$ | T |
| $\operatorname{Var}(\mathbf{Y}) = \sigma^2\mathbf{I}_5$ | T |
| $\mathbb{E}[\mathbf{Y}] = \mathbf{0}$ | F |
| The five errors all have variance $\sigma^2$ | T |

:::

:::

::::

## Question 2: Least Squares and Projection [12]

Continue using the data and design matrix from Question 1.

### (a) [4]

Starting from

$$
S(\boldsymbol{\beta})
=
(\mathbf{Y}-\mathbf{X}\boldsymbol{\beta})^\top
(\mathbf{Y}-\mathbf{X}\boldsymbol{\beta}),
$$

derive the normal equations

$$
\mathbf{X}^\top\mathbf{X}\hat{\boldsymbol{\beta}}
=
\mathbf{X}^\top\mathbf{Y},
$$

and then state the OLS estimator when $\mathbf{X}$ has full column rank.

Keep your derivation concise.

:::: {.content-visible when-meta="show-solutions"}

::: {#a1-solution-2a .callout-tip collapse="true" title="Solution 2(a)"}

Expanding,

$$
S(\boldsymbol{\beta})
=
\mathbf{Y}^\top\mathbf{Y}
-
2\boldsymbol{\beta}^\top\mathbf{X}^\top\mathbf{Y}
+
\boldsymbol{\beta}^\top\mathbf{X}^\top\mathbf{X}\boldsymbol{\beta}.
$$

Differentiating,

$$
\frac{\partial S(\boldsymbol{\beta})}{\partial\boldsymbol{\beta}}
=
-2\mathbf{X}^\top\mathbf{Y}
+
2\mathbf{X}^\top\mathbf{X}\boldsymbol{\beta}.
$$

Setting the derivative equal to zero gives

$$
\mathbf{X}^\top\mathbf{X}\hat{\boldsymbol{\beta}}
=
\mathbf{X}^\top\mathbf{Y}.
$$

Hence,

$$
\boxed{
\hat{\boldsymbol{\beta}}
=
(\mathbf{X}^\top\mathbf{X})^{-1}
\mathbf{X}^\top\mathbf{Y}
}.
$$

:::

::::

### (b) [3]

Using the quantities from Question 1, compute

$$
\hat{\boldsymbol{\beta}}
=
\begin{pmatrix}
\hat{\beta}_0\\
\hat{\beta}_1
\end{pmatrix}.
$$

Fill in the blanks:

$$
\hat{\beta}_0 = \underline{\hspace{2cm}},
\qquad
\hat{\beta}_1 = \underline{\hspace{2cm}}.
$$

Hence,

$$
\hat{Y}
=
\underline{\hspace{2cm}}
+
\underline{\hspace{2cm}}x.
$$

:::: {.content-visible when-meta="show-solutions"}

::: {#a1-solution-2b .callout-tip collapse="true" title="Solution 2(b)"}

Using the data in Question 1,

$$
\hat{\boldsymbol{\beta}}
=\frac{1}{50}\begin{pmatrix}30&-10\\-10&5\end{pmatrix}
\begin{pmatrix}28\\74\end{pmatrix}
=\begin{pmatrix}2.0\\1.8\end{pmatrix}.
$$

Therefore $\hat{\beta}_0=2.000$, $\hat{\beta}_1=1.800$, and

$$
\boxed{\hat Y=2.0+1.8x}.
$$

These are the coefficients used in Question 3.

:::

::::

### (c) [2]

For the least squares residual vector

$$
\mathbf{e}
=
\mathbf{Y}-\hat{\mathbf{Y}},
$$

which statement must be true?

A.  $\mathbf{X}\mathbf{e}=\mathbf{0}$

B.  $\mathbf{X}^\top\mathbf{e}=\mathbf{0}$

C.  $\mathbf{e}^\top\mathbf{e}=0$

D.  $\mathbf{H}\mathbf{e}=\mathbf{e}$

In one sentence, explain the geometric meaning of your choice.

**Your explanation:**  
`____________________________________________________________`

:::: {.content-visible when-meta="show-solutions"}

::: {#a1-solution-2c .callout-tip collapse="true" title="Solution 2(c)"}

**B.** The residual vector is orthogonal to the column space of $\mathbf{X}$.

:::

::::

### (d) [3]

For the hat matrix

$$
\mathbf{H}
=
\mathbf{X}
(\mathbf{X}^\top\mathbf{X})^{-1}
\mathbf{X}^\top,
$$

enter `T` or `F`.

::: {tbl-colwidths="[88,12]"}

| Statement | T/F |
|:---|:---:|
| $\mathbf{H}^\top=\mathbf{H}$ | `___` |
| $\mathbf{H}^2=\mathbf{H}$ | `___` |
| $\hat{\mathbf{Y}}=\mathbf{H}\mathbf{Y}$ | `___` |
| $\mathbf{H}$ is an orthogonal projection matrix | `___` |

:::

:::: {.content-visible when-meta="show-solutions"}

::: {#a1-solution-2d .callout-tip collapse="true" title="Solution 2(d)"}

::: {tbl-colwidths="[88,12]"}

| Statement | T/F |
|:---|:---:|
| $\mathbf{H}^\top=\mathbf{H}$ | T |
| $\mathbf{H}^2=\mathbf{H}$ | T |
| $\hat{\mathbf{Y}}=\mathbf{H}\mathbf{Y}$ | T |
| $\mathbf{H}$ is an orthogonal projection matrix | T |

:::

:::

::::

## Question 3: Consequences and Geometry of Least Squares [12]

Continue using the data and fitted regression model from Questions 1 and 2.

Recall that

$$
\hat{\beta}_0=2.0,
\qquad
\hat{\beta}_1=1.8.
$$

### (a) Fitted Values and Residuals [3]

Compute the fitted values

$$
\hat{\mathbf{Y}}
=
\mathbf{X}\hat{\boldsymbol{\beta}}
$$

and the residual vector

$$
\mathbf{e}
=
\mathbf{Y}-\hat{\mathbf{Y}}.
$$

Fill in the blanks:

$$
\hat{\mathbf{Y}}
=
\begin{pmatrix}
\underline{\hspace{1cm}}\\
\underline{\hspace{1cm}}\\
\underline{\hspace{1cm}}\\
\underline{\hspace{1cm}}\\
\underline{\hspace{1cm}}
\end{pmatrix},
\qquad
\mathbf{e}
=
\begin{pmatrix}
\underline{\hspace{1cm}}\\
\underline{\hspace{1cm}}\\
\underline{\hspace{1cm}}\\
\underline{\hspace{1cm}}\\
\underline{\hspace{1cm}}
\end{pmatrix}.
$$

:::: {.content-visible when-meta="show-solutions"}

::: {#a1-solution-3a .callout-tip collapse="true" title="Solution 3(a)"}

$$
\hat{\mathbf{Y}}
=
\begin{pmatrix}
2.0\\
3.8\\
5.6\\
7.4\\
9.2
\end{pmatrix},
\qquad
\mathbf{e}
=
\begin{pmatrix}
0\\
0.2\\
-0.6\\
0.6\\
-0.2
\end{pmatrix}.
$$

:::

::::

### (b) Consequences of Including an Intercept [3]

For each statement, enter `T` or `F`.

::: {tbl-colwidths="[88,12]"}

| Statement | T/F |
| :--- | :---: |
| The residuals satisfy $\sum_{i=1}^n e_i=0$. | `___` |
| The fitted values satisfy $\bar{\hat{Y}}=\bar{Y}$. | `___` |
| The vector $\mathbf{1}_n$ belongs to $\mathcal{C}(\mathbf{X})$. | `___` |
| These properties necessarily hold for every regression model without an intercept. | `___` |

:::

:::: {.content-visible when-meta="show-solutions"}

::: {#a1-solution-3b .callout-tip collapse="true" title="Solution 3(b)"}

::: {tbl-colwidths="[88,12]"}

| Statement | T/F |
| :--- | :---: |
| The residuals satisfy $\sum_{i=1}^n e_i=0$. | T |
| The fitted values satisfy $\bar{\hat{Y}}=\bar{Y}$. | T |
| The vector $\mathbf{1}_n$ belongs to $\mathcal{C}(\mathbf{X})$. | T |
| These properties necessarily hold for every regression model without an intercept. | F |

:::

Because the model contains an intercept, $\mathbf{1}_n$ is a column of
$\mathbf{X}$. Since $\mathbf{X}^\top\mathbf{e}=\mathbf{0}$,

$$
\mathbf{1}_n^\top\mathbf{e}=0,
$$

which implies

$$
\sum_{i=1}^n e_i=0
$$

and therefore

$$
\bar{\hat{Y}}=\bar{Y}.
$$

:::

::::

### (c) Residual-Maker Matrix [3]

Define

$$
\mathbf{M}
=
\mathbf{I}_n-\mathbf{H}.
$$

For each statement, enter `T` or `F`.

::: {tbl-colwidths="[88,12]"}

| Statement | T/F |
| :--- | :---: |
| $\mathbf{e}=\mathbf{M}\mathbf{Y}$ | `___` |
| $\mathbf{M}^\top=\mathbf{M}$ | `___` |
| $\mathbf{M}^2=\mathbf{M}$ | `___` |
| $\mathbf{H}\mathbf{M}=\mathbf{0}$ | `___` |

:::

:::: {.content-visible when-meta="show-solutions"}

::: {#a1-solution-3c .callout-tip collapse="true" title="Solution 3(c)"}

::: {tbl-colwidths="[88,12]"}

| Statement | T/F |
| :--- | :---: |
| $\mathbf{e}=\mathbf{M}\mathbf{Y}$ | T |
| $\mathbf{M}^\top=\mathbf{M}$ | T |
| $\mathbf{M}^2=\mathbf{M}$ | T |
| $\mathbf{H}\mathbf{M}=\mathbf{0}$ | T |

:::

The matrix $\mathbf{H}$ projects onto $\mathcal{C}(\mathbf{X})$,
whereas $\mathbf{M}$ projects onto
$\mathcal{C}(\mathbf{X})^\perp$.

:::

::::

### (d) Rank and Uniqueness [3]

Suppose now that a design matrix $\mathbf{X}$ does **not** have full
column rank.

For each statement, enter `T` or `F`.

::: {tbl-colwidths="[88,12]"}

| Statement | T/F |
| :--- | :---: |
| $(\mathbf{X}^\top\mathbf{X})^{-1}$ necessarily exists. | `___` |
| The least squares coefficient vector $\hat{\boldsymbol{\beta}}$ may not be unique. | `___` |
| The fitted vector $\hat{\mathbf{Y}}$ is still unique. | `___` |
| The residual vector $\mathbf{e}$ is still unique. | `___` |

:::

:::: {.content-visible when-meta="show-solutions"}

::: {#a1-solution-3d .callout-tip collapse="true" title="Solution 3(d)"}

::: {tbl-colwidths="[88,12]"}

| Statement | T/F |
| :--- | :---: |
| $(\mathbf{X}^\top\mathbf{X})^{-1}$ necessarily exists. | F |
| The least squares coefficient vector $\hat{\boldsymbol{\beta}}$ may not be unique. | T |
| The fitted vector $\hat{\mathbf{Y}}$ is still unique. | T |
| The residual vector $\mathbf{e}$ is still unique. | T |

:::

Even when several coefficient vectors produce the same least squares
solution, the orthogonal projection of $\mathbf{Y}$ onto
$\mathcal{C}(\mathbf{X})$ is unique.

:::

::::

## Question 4: Linear Regression in R [8]

For this question, use the built-in R dataset `cars`.

The variables are:

- `speed`: speed of the car in miles per hour;
- `dist`: stopping distance in feet.

Consider the model

$$
\mathrm{dist}_i
=
\beta_0
+
\beta_1\mathrm{speed}_i
+
\varepsilon_i.
$$

Load the data using

```r
data(cars)
```

### (a) [4]

Fit the model using `lm()` and complete the table.

```r
fit <- lm(dist ~ speed, data = cars)
summary(fit)
```

| Quantity | Your answer |
|:---|:---|
| $\hat{\beta}_0$ | `___` |
| $\hat{\beta}_1$ | `___` |

Which interpretation of $\hat{\beta}_1$ is correct?

A.  For each 1 mph increase in speed, the stopping distance increases by exactly 3.9324 feet for every car.

B.  For each 1 mph increase in speed, the fitted mean stopping distance increases by approximately 3.9324 feet.

C.  The average speed increases by 3.9324 mph for every extra foot of stopping distance.

D.  A car travelling at 0 mph has a stopping distance of 3.9324 feet.

:::: {.content-visible when-meta="show-solutions"}

::: {#a1-solution-4a .callout-tip collapse="true" title="Solution 4(a)"}

```{r}
#| code-fold: false
data(cars)
fit <- lm(dist ~ speed, data = cars)
coef(fit)
```

$$
\hat{\beta}_0=-17.5791,
\qquad
\hat{\beta}_1=3.9324.
$$

The correct interpretation is **B**.

:::

::::

### (b) [4]

Construct $\mathbf{Y}$ and $\mathbf{X}$ manually and reproduce the OLS coefficients using matrix operations.

Complete the following code:

```r
Y <- matrix(cars$dist, ncol = 1)

X <- cbind(
  1,
  cars$speed
)

# Replace NULL with your matrix expression
beta_hat <- NULL

beta_hat
```

Then report

$$
\hat{\boldsymbol{\beta}}
=
\begin{pmatrix}
\underline{\hspace{2cm}}\\
\underline{\hspace{2cm}}
\end{pmatrix}.
$$

:::: {.content-visible when-meta="show-solutions"}

::: {#a1-solution-4b .callout-tip collapse="true" title="Solution 4(b)"}

```{r}
#| code-fold: false
Y <- matrix(cars$dist, ncol = 1)
X <- cbind(1, cars$speed)
beta_hat <- solve(t(X) %*% X) %*% t(X) %*% Y
beta_hat
```

The result agrees with `lm()`:

$$
\hat{\boldsymbol{\beta}}
=\begin{pmatrix}-17.579095\\3.932409\end{pmatrix}
\approx\begin{pmatrix}-17.579\\3.932\end{pmatrix}.
$$

:::

::::
